Credit risk distribution and capital savings

### 2019-08-14 Analysis/Descriptive/ExtendedAbstract/ExtendedAbstract.Rmd
Update extended abstract before submitting to Copfir conference

---
output: 
  pdf_document:
    citation_package: natbib
    latex_engine: pdflatex
    template: tex-ms.tex
title: "Credit risk distribution and capital savings"
subtitle: ""
thanks: ""
date: # Today, unless specified. In smaller font if specified to allow for version numbering
author:
- name: Lucas Avezum
  affiliation: Tilburg University; l.avezum@uvt.nl; +31134664531;  Warandelaan 2, 5037 AB Tilburg, NL, The Netherlands

@@

This paper shows that higher-order moments of the distribution of banks' portfolio credit risk, in addition to its average, are important factors of risk-weights (RWs) variability across banks. More specifically, two distributions of probabilities of default (PD) with the same average but different variance result in different capital charges. Using a novel hand-collected dataset, I propose a measure that disantangles this PD distribution effect from the portfolio's average contribution to capital requirements. I estimate the impact of this measure on banks' performance.

Under the Internal ratings-based approach (IRB) banks supply their own risk estimates to regulatory formulas which determine the amount of capital banks have to hold for each asset. Importantly, these formulas are concave. This means, by Jensen’s inequality, that the sum of capital charges for two assets is lower than the charge for their average (in terms of PD). More generally, the distribution of credit risk matters to calculate capital requirements with, for instance, disperse portfolios yielding relatively lower capital charges.

@@

I then extend the analysis to find out whether banks adjust their portfolio risk distribution in response to capital requirement shocks. In the spirit of @gropp2018banks, I explore the 2011 EBA capital exercise as a quasi-natural experiment. This was a strong and unexpected test to the solvency of 63 European banks that resulted in a recommendation for these banks to build up their capital buffer to reach a 9.0% core tier 1 capital ratio by June 2012. Identification of the effect of higher capital requirements on banks’ portfolio risk distribution comes from correlating changes in the _capital savings_ measure around the exercise to the distance to achieve the 9.0% required ratio for banks that participated in the exercise. Once established that banks adjust their portfolio risk distribution to achieve regulatory capital requirements, a natural extension is to quantify the effect of these changes on capital ratio and evaluate if these changes impact other variables of interest, such as, return on equity, and risk-taking. To solve endogeneity concerns – for instance, worse capitalized banks have stronger incentives to adjust their portfolios – I use the Basel 2 adoption year across countries as an instrument for the PD distribution effect.

Taken together, the results show that strategic portfolio choice, in terms of credit risk, is an important way banks can reduce capital charges while maintaining their average portfolio risk and return on equity.

\newpage
\singlespacing

@@

>>>>

---
output: 
  pdf_document:
    citation_package: natbib
    latex_engine: pdflatex
    template: tex-ms.tex
title: "Credit risk distribution and capital savings"
subtitle: ""
thanks: ""
date: # Today, unless specified. In smaller font if specified to allow for version numbering
author:
- name: Lucas Avezum
  affiliation: CentER and EBC, Tilburg University; l.avezum@uvt.nl; +31134664531;  Warandelaan 2, 5037 AB Tilburg, NL, The Netherlands

@@

This paper shows that higher-order moments of the distribution of banks' portfolio credit risk, in addition to its average, are important factors of risk-weights (RWs) variability across banks. More specifically, two distributions of probabilities of default (PD) with the same average but different variance result in different capital charges. Using a novel hand-collected dataset, I propose a measure that disantangles this PD distribution effect from the portfolio's average contribution to capital requirements. I then estimate the impact of this measure on banks' performance.

Several studies document significant RW variability across otherwise similar banks. Many take this evidence as indicative of either excessive subjectivity in the current capital requirement rules or risk measurement manipulation by banks to decrease capital charges [see,  @mariathasan2014manipulation; @ferribank]. Others are more cautious and emphasize that part of RWA variability is desirable, since it reflects differences in risk among banks [see, @cannata2012inside;@BCBS2016regulatory; @EBA2019result]. I contribute to the debate by showing that the PD distribution effect is an important factor of the RW variation across banks. This implies that previous studies in the literature are potentially overestimating subjectivity and manipulation of RWs. Nonetheless, the PD distribution effect entails unintended effects to banks’ solvency as they can explore this feature to reduce capital requirements without altering their portfolio average risk.

Under the internal ratings-based approach (IRB) banks supply their own risk estimates to regulatory formulas which determine the amount of capital banks have to hold for each asset. Importantly, these formulas are concave. This means, by Jensen’s inequality, that the sum of capital charges for two assets is lower than the charge for their average (in terms of PD). More generally, the distribution of credit risk matters to calculate capital requirements with, for instance, disperse portfolios yielding relatively lower capital charges.

@@

I then extend the analysis to find out whether banks adjust their portfolio risk distribution in response to capital requirement shocks. In the spirit of @gropp2018banks, I explore the 2011 EBA capital exercise as a quasi-natural experiment. This was a strong and unexpected test to the solvency of 63 European banks that resulted in a recommendation for these banks to build up their capital buffer to reach a 9.0% core tier 1 capital ratio by June 2012. Identification of the effect of higher capital requirements on banks’ portfolio risk distribution comes from correlating changes in the PD distribution effect measure around the exercise to the distance to achieve the 9.0% required ratio for banks that participated in the exercise. Once established that banks adjust their portfolio risk distribution to achieve regulatory capital requirements, a natural extension is to quantify the effect of these changes on capital ratio and evaluate if these changes impact other variables of interest, such as, return on equity, and risk-taking. To alleviate endogeneity concerns – for instance, worse capitalized banks have stronger incentives to adjust their portfolios – I use the Basel 2 adoption year across countries as an instrument for the PD distribution effect.

Taken together, the results show that strategic portfolio choice, in terms of credit risk, is an important way banks can reduce capital charges while maintaining their average portfolio risk and return on equity. Differently from manipulation and subjectivity of RW under IRB, the PD distribution effect hereby identified does not entail any regulatory breach. However, the assumption that riskier assets have lower systematic risk, which is one of the reasons
the regulatory formulas are concave, may be too simplistic to capture the true systematic component of each asset. If that is the case, this paper shows that the current capital requirements framework does not correctly equate capital charges to credit risk, consequently, undermining financial stability.

\singlespacing

@@

\newpage
